Scaling up a GF grammar for free parsing

### 2011-12-11 doc/report/report.tex

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{lingmacros}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{wrapfig}
% exempel: 2) Inte vet jag
%              not know I
%             I don't know

@@

\tableofcontents
\newpage

\abstract{
% This is the report of the work of expanding a GF grammar and preparing
% it for parsing of free Swedish. A lexicon consisting of more than 100 000 words
% has been extracted and we have constructed an automatic mapping between the
% annotation of the treebank Talbanken and GF. 
% The goal is to build a robust parser dealing with uncontrolled natural
% language, a parser which, for example, can parse all of Talbanken.
% Outside of this project, an equivalent GF project for English is being worked upon,
% as are theqniques for making the parser robust.
% With this in mind, we have extended the current GF grammar to adapt it for
% Swedish specifically so that it can parse the an extended set of basic
% constructions. We have completed a method for importing the eletronical 
% lexicon Saldo to a GF format and upprätat en connection mellan oss och Talbanken.

@@

% check references! lncs is good maybe. lnai
\newpage
\chapter{Introduction}
%\section{Introduction} 
The grammar formalism GF has so far been succesfull used for 
controlled languges, but recent experiments have showed
that it is also possible to use the framework for parsing free language.
Parsing, that is the task of automatically identifiyng morpohlogical and
syntactical structures to text, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available ruled-based parser that gives a deep analyse for
Swedish.  There are other parsers, but the fact that they are not freely available does not only
restrict the usage but also their possiblities of being developed. Our goal is
to create a open source Swedish grammar and parser that can be continuosly enhanced and
be made use of in other projects.
To build a parser or a grammar from scratch is a heavy work and we
therefore combine already existing parts.
We utgår från GF resource grammar and therefore have 
The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other langugages implentented in the framework.
The grammar defines a parser, which accepts all constructions defined by the grammar.

We hence needed to extend the Swedish grammar to cover
the language specific constructions. Further a large lexicon needed, as well
as a method for evaluation.
During this project, we have worked on and explored those things and some problems identified.
%We have extended the Swedish resource grammar in GF, to give a better coverage and imported
%a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

@@

\section{Background}  
The work described in this is thesis is part of a bigger project which aims at
using GF for parsing free Swedish. 
In previous work \ref{??} the start on extending the Swedish GF grammar was
made and a tool for lexical aquistion were developed.
We now construct a bigger
and more expressive grammar as well as a large scale GF lexicon.
As all GF grammars, this one can be used for parsing, and we develop it by getting
examples, ideas and test material from a treebank.
The project is hence heavily depending on three resources, which will be described
in this section.

\subsection{Swedish}
Swedish\cite[Inl. \textsection 3]{SAG}is a North-Germanic language,
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
Swedish syntax are often similar to English, but have some characteristics:
\subsubsection*{Verb second}
Swedish is a verb-second language\cite[p. 116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be
fronted\cite[\textsection 1027]{H&H}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being emphasized.
\enumsentence{
\shortex{4}
{Du & ser & mig & inte.}
{You & see & me & not.}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{Me & see & you & not.}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{Not&  see& you& me.}} \label{ex:swedish-avso}

%\shortexm{3}
%{hej} {hi}{hhe} {bing}{bong}{salt}
%; adverbials, objects
%Alla syntaktiska kategorier kan be fronted utom obetonade satsadverbial (ju,väl).
%The first constituent of the clause is usually made up of the subject,
%although it likewise could consist of adverbial phrases or objects that are
%being empathazied. \\
%\enumsentence{
%\begin{tabular}{lll}
%  Johan & gick & sakta på gatan.\\
%\emph{Johan} & \emph{walked} & \emph{slowely on the street.}\\
%  Sakta & gick & Johan på gatan.\\
%\emph{Slowly} & \emph{walked} & \emph{Johan on the street.}\\
%  På gatan & gick & Johan sakta.\\
%\emph{On the street} & \emph{walked}  & \emph{Johan slowly.}\\
%\end{tabular}}

Inverted word order marks questions
\enumsentence{
\shortex{4}
{Såg & du & mig & inte?}
{Saw  & you&  me& not?}
{`Didn't you see me?'}}
%\enumsentence{Gick Johan på gatan?\\
%        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
>> and stricter
%\enumsentence{
%\begin{tabular}{ll}
%{\shortex{7}
%{Jag & såg&  Anna.&  Hon & såg & inte&  mig.} 
%{ & & &  & & &}}
%\shortex{7}
%{Jag såg att Anna \textbf{inte} såg mig.}\\
%\shortex{7}
%{I  saw Anna. She saw not   me.}  & {I   saw that Anna not saw me}. \\
%`I saw Anna. $\;$ She did not see me.'& `I saw that Anna did not see me'.
%}
\enumsentence{Main sentences:\\
{\shortex{7}{Jag & såg & Anna. & Hon & såg & {\bf inte} & mig }
{ I & saw & Anna. & She & saw & {\bf not} & me}
{`I saw Anna. She didn't see me'}}}
\enumsentence{Subordinate sentence:\\
\shortex{7}{Jag & såg & att& Anna& \textbf{inte}& såg& mig.}
            {I &  saw & that& Anna& \textbf{not}& saw& me.}
            {`I saw that Anna didn't see me'}}

\subsubsection*{Passive voice}
Passive voice is often used in Swedish, especially the
\textit{s-passive}, which is formed by adding an \emph{s} at the end of a verb.
\enumsentence{Passive \hspace{39mm} Active\\
\shortexm{11}{Han & jagade\textbf{s} & av & ett & lejon.}
        {He & hunted+\textbf{s} & by & a & lion.}
        {`He was hunted by a lion.'}
{Ett & lejon & jagade & honom.}
        {A &lion &hunted &him.}
        {`A lion was hunting him.'}}

%cf.
%\enumsentence{
%\shortex{5}{Ett & lejon & jagade & honom.}
%        {A &lion &hunted &him.}
%        {`A lion was hunting him.'}}

\subsubsection*{Impersonal constructions}
Constructions with 'det är/var' ('it is/was') is very common\cite[\textsection 309d]{H&H}:
\enumsentence{
\shortex{6}{Det & var & roligt & att & höra.}
{It &was & nice &to &hear.}
{`I'm glad to hear that.'}}
'Det' (\emph{it}) is also used as formal subject %for intransitive verbs,
with the real subject put in the position of an object.
\enumsentence{
\shortex{6}{Det &står &en &älg &på &fältet.}
{ It & stands &a & moose &on &the field.}
{`There is a moose on the field.'}}

\subsubsection*{Reflexive pronouns}
\label{swe:refl}
The Scandinavian language have special reflexive pronouns
and possessive pronouns for the 3rd
person\cite[\textsection 310 \& 319]{H&H}. Those are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{ll}
Han slog \textbf{sig}. & Han såg \textbf{sitt} barn.\\
\vspace{3mm}
\emph{He hit him self.} & \emph{He saw his (own) child.}\\
Han slog \textbf{honom}. & Han såg \textbf{hans} barn.\\
\emph{He hit him (another person).}&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form.

Mention varandra. Same idea.
%Basic info about Swedish. V2 lang, inverted word order and subordinate clauses.
%Passive, reflexive 'sitt', 'det är', 'det sitter en katt där', prepare the reader
%for what will be written in the grammar part.
%Adjectives but not verbs congruate with nouns. De är stora, They are big, Jag
%är stor, I am big
%Since the s-passive is the most common one in written Swedish \cite{SAG-34-1}, it has

\subsection{Saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The words are equipped with semantic, syntactic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\newpage % only until saldo can be on same page as the pic
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

If one word
have different meanings, there will be separate entries for each of them.
The word ``växa" (``grow") for example, has two meanings: \emph{getting bigger}
\enumsentence{Gräset växte varje dag\\
The grass grew each day}
or \emph{being a plant}. 
\enumsentence{Det växte en gammal ek där.\\
There stood an old oak.}
The word ``växa" is however declined the same way for both interpretations of the word.
Hence Saldo's morphological lexicon, provided under LPGL in XML format,
only keeps one entry of ``växa".
For the homonyms that does not share the whole inflection tables, different entries are 
kept in the morphological lexicon.

@@

\subsection{Grammatical Framework}
>> extensnion to CFG, same expressiveness as TAG and HG. parse in polynomial time.
>> incremental parsing.
>> how parse trees work, not correct complete representation.
>> not a constituent structure. compare to lfg, fstructure, cstructure.
>> explain how they are created, graphviz and how a user uses them.
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a grammar formalism based on Martin-Löf type theory\cite{martinlof}. GF is
designed for multilingual applications and the formalism is equivalent to
Parallel Multiple Context-Free Grammars (Seki \& al. 1991), it is hence
stronger than mildly contex-free grammars.

GF is a functional programming languages, inspired by programming languages
like ML\cite{ml}, Haskell\cite{haskell}, Prolog\cite{bratko} and
Agda\cite{agda}. 
>> GF ~ 15 years. PGF= runtime systm
It has been developed for about 20 years 
>> snd: functional and programming language, type checking and logic: prolog (check this)
>> dependent and implicit types : agda. (read gf book and krasses book)

A grammar written in GF can be used for both parsing and generation.
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system. The interpreter, the GF shell, lets
you try and test your grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
The compiler lets you compile your grammar into portable grammar format (PGF)\ref{pgf}.
There are libraries for using PGF files with Java, Java script and Haskell. The grammars
>> Haskell only complete, also or C\# and C. 
may also be used in Android applications.
and are inspired by.\\

The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
>> internal structure
>> compare fstructure in LFG
shared by many languages and can be implemented by a numeber of concrete grammars,
each representing one language. The abstract syntax models the semantics of the
domain without concern for language specific features such as agreement or word
order.

@@

Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|,
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. To do this, we add a category \verb|NP| to the
abstract. \verb|N| now means a noun
in any of the number, and we introduce two function, \verb|NSg : N -> NP| and
\verb|NPl : N -> NP| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; NP ; V ; S ;               lincat V, S, NP = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : NP -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> NP ;                    NPl n = n.s ! Pl ;
    NPl : N -> NP ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records,tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s| shows that it is a table, which given a parameter of type \verb|Num| returns
a string. \verb|Num| is defined as a parameter, which can either have value \verb|Sg|
or \verb|Pl|. In \verb|NPl| and \verb|NSg|, \verb|n.s| means that we use the
field \verb|s| of \verb|n|, and then the selection operator \verb|!| is used to
select a branch in the table.

@@

>>not very restricted, not few ambiguities.
>>the scope is controlled language, but possible ambiguities (discuss this in future
>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
>>syntactical (man with the telescope), 
>>lexical (youSg/Pl, växa).
A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt},
a project which aims to develop language independent material for mathematics problems,
and the formal software verification tool KeY\cite{key}, in wich GF is used for
translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
European project MOLTO\cite{molto} for online translation between up to 15 language.\\

\subsubsection{The resource library}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a powerful resource library\cite{gf-resource}, which covers the
fundamental morphology and syntactics for more than 20 languages. %is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and they constitute  
a very useful resource when implementing application grammars.
There is also a small test lexicon included, containing some hundred of common
words.

>>>>

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{lingmacros}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{wrapfig}
\usepackage{tipa}
% exempel: 2) Inte vet jag
%              not know I
%             I don't know

@@

\abstract{
This is the report of the work of expanding and enhancing a GF grammar and preparing
it for parsing of free Swedish. 
Outside of this project, an equivalent project for English is being worked upon,
as are tehcniques for making the parser robust.
With this in mind, we have extended the current GF grammar to adapt it for
Swedish specifically so that it can parse an extended set of basic
constructions. We have completed a method for importing the electronical 
lexicon Saldo to a GF format and established a connection to the treebank Talbanken.
The imported lexicon consists of more than 100 000 words
and the grammar now covers ??\% of the xx part of Talbanken. 
% The mapping between the annotations of Talbanken and GF is automatic. 
% The goal is to build a robust parser dealing with uncontrolled natural
%language, a parser which can parse all of Talbanken.

@@

% check references! lncs is good maybe. lnai
\newpage

\tableofcontents
\newpage
\chapter{Introduction}
>> short intro to GF
The grammar formalism GF has so far been succesfull used for 
controlled languges, but recent experiments have showed
that it is also possible to use the framework for parsing free language.
Parsing, that is the task of automatically identifiyng morpohlogical and
syntactical structures to text, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that give a deep analyse for
Swedish. The fact that the existing rule-based Swedish
parsers are not freely available does not only restrict their
usage but also their possiblities of being developed. Our
goal is
to create a open source Swedish grammar and parser that can be continuosly enhanced and
be made use of in other projects.\\
To build a parser or a grammar from scratch is a heavy work ~~ and we
therefore combine already existing parts of ??.
We start from GF resource grammar and thereby get
%The usage of GF allows us to start from
a well-defined system for describing grammar, as well as a strong connection to
more than 20 other langugages implentented in the framework.
>>The grammar defines a parser, which accepts all constructions defined by the grammar.

\section{Aims}
>> enhance gf...
The purpose of this project has been to prepare the GF grammar for free parsing.
This has meant to develop earlier techniques to fit for Swedish, create methods
for achieving and keeping a extensive lexicon and to adapt the existing
resource grammar to fit for more language specific constrictions.
We hence needed to extend the Swedish grammar to cover
the language specific constructions. Further a large lexicon was
needed, as well as a method for evaluation.
During this project, we have worked on and explored those things and
while identifying problems and future work.

\section{Outline}
This thesis is divided into 6 chapters. We start by giving some background
information to areas relevant to the project. %Chapter \ref{sec:background} 
Grammatical Framework is presented in section \ref{sec:gf} and a deeper
presentation of the Swedish resource grammar is given in section \ref{sec:swegf}.
Section \ref{sec:swedish} gives an introduction Swedish and
brief presentations of Saldo and Talbanken are found in section \ref{sec:saldo} and
\ref{sec:talbanken} respectively.
A summary of related work can be found in section \ref{sec:related}.

Chapters \ref{sec:prog.saldo} - \ref{sec:Mapping} present the methodology,
implementation and some results. Finally, conclusions, an evaluation and a list
of future work are presented in chapter \ref{sec:results}.

%We have extended the Swedish resource grammar in GF, to give a better coverage and imported
%a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

@@

\chapter{Background}  
\label{sec:background}
The work described in this is thesis is part of a bigger project which aims at
using GF for parsing free Swedish. 
In previous work \ref{??} the start on extending the Swedish GF grammar was
made and a tool for lexical aquistion were developed.
We now construct a bigger
and more expressive grammar as well as a large scale GF lexicon.
As all GF grammars, this one can be used for parsing, and we develop it by getting
examples, ideas and test material from a treebank.
The project is hence heavily depending on three resources, which will be described
in this section.

\section{Swedish}
\label{sec:swedish}
>> move, have GF first?\\
Swedish\cite[Inl. \textsection 3]{SAG}is a North-Germanic language,
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutually intelligible. Swedish is also 
an official language in Finland and altogether spoken by approximately 10 ~~
9?Check SAG million people.
Swedish syntax is often similar to English, but have some characteristics ~~:

\subsection*{Post-nominal articles} 
Unlike most of the world's languages Swedish express not only the number but also
the definitness of a noun by suffixes. The endings are decided by the gender of the noun,
which may be \textit{neuter} or \textit{common}.\\
%(or \textit{utrum})

\begin{tabular}{l|lll}
&\textbf{Definite}& \textbf{Indefinite}& \textbf{Gender} \\
 \hline
\textbf{Singular} & {en katt} & {katten} & \textit{common}\\
& \emph{a cat} & \emph{the cat} &\\
\textbf{Plural} & katter & katterna &\\
 &\emph{cats} & \emph{the cats} &\\
 \hline
\textbf{Singular} & {ett hus} & {huset} & \textit{neuter}\\
 & \emph{a house} & \emph{the house} & \\
\textbf{Plural} & hus & husen & \\
& \emph{houses} & \emph{the houses} & \\
\end{tabular}\\

The definiteness can be accentuated by using the definite articles.\\

\begin{tabular}{llll}
den katten & det huset & de katterna & de husen \\
\emph{the cat} & \emph{the house} & \emph{the cats} & \emph{the houses} \\
\end{tabular}\\

The definite articles are required when the noun is modified by adjectival phrase, sentence
\ref{sent:trottkatt}b is hence not grammatically correct.
%\enumsentence{\shortexnt{4}
%{Den& trötta& katten & sov}
%{the& tired& cat+\textbf{the} & slept}}
\enumsentence{\begin{tabular}{llllll}
a.& Katten sov. & b.& *Trötta katten sov &c. &Den trötta katten sov\\
&The cat slept. & &~~The tired cat slept& &The tired cat slept\\
\end{tabular}}\label{sent:trottkatt}

%can be superdoubled
%example: den där boken

\subsection*{Verb second}
Swedish is a verb-second language\cite[p.116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be
fronted\cite[\textsection 1027]{H&H}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, but vary in how the content is presented.
%~~ although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being
%emphasized.
\enumsentence{
\shortex{4}
{Du & ser &  inte &mig.}
%empha glossning!! ha samma överallt, ' eller ''
{you & see & not & me}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{me & see & you & not}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{not&  see& you& me}} \label{ex:swedish-avso}

%Alla syntaktiska kategorier kan be fronted utom obetonade satsadverbial (ju,väl).
%\enumsentence{
%\begin{tabular}{lll}
%  Johan & gick & sakta på gatan.\\
%\emph{Johan} & \emph{walked} & \emph{slowely on the street.}\\
%  Sakta & gick & Johan på gatan.\\
%\emph{Slowly} & \emph{walked} & \emph{Johan on the street.}\\
%  På gatan & gick & Johan sakta.\\
%\emph{On the street} & \emph{walked}  & \emph{Johan slowly.}\\
%\end{tabular}}

Inverted word order marks questions
\enumsentence{
\shortex{4}
{Såg & du & mig & inte?}
{saw  & you&  me& not}
{`Didn't you see me?'}}
%\enumsentence{Gick Johan på gatan?\\
%        \emph{Did Johan walk on the street?}}
The word order in subordinate clauses in also slightly different,
%>> and stricter
\enumsentence{Main sentences:\\
{\shortex{8}{Jag & såg & Anna & men & hon & såg & {\bf inte} & mig }
{ I & saw & Anna & but & she & saw & {\bf not} & me}
{`I saw Anna but she didn't see me'}}}
\enumsentence{Subordinate sentence:\\
\shortex{7}{Jag & förstod & att& Anna& \textbf{inte}& såg& mig.}
            {I &  understood & that& Anna& \textbf{not}& saw& me}
            {`I understood that Anna didn't see me'}}

\subsection*{Passive voice}
There are two ways of forming passive verb phrases in Swedish: the 
\textbf{periphrastic passive}, formed by using a modale auxiliary verb \\
%>> better with animate subjects
\enumsentence{
\shortex{5}
{Tjuven &blev& tagen& av & polisen}
{the thief & was & taken & by & the police }
{`The thief was arrested by the police'}}
\label{gfPass:peri-pass}
and the \textbf{s-passive} which is formed by adding an \emph{s} to the verb: \\
\enumsentence{Passive \hspace{45mm} Active\\
\shortexm{12}
{Tjuven & togs&  av & polisen}
{the thief&  took+\textbf{s}&  by&the police}
{`The thief was arrested by the police'}
{&&Polisen & tog & tjuven}
{&&the police & took & the thief}
{\hspace{-9mm}`The police arrested the thief'}}
\label{gfPass:s-pass}
The s-passive is the most commonly used one for both written
and Swedish \cite[??]{SAG} and dominates especially when the subject is inanimate.
%cf.
%\enumsentence{
%\shortex{5}{Ett & lejon & jagade & honom.}
%        {A &lion &hunted &him.}
%        {`A lion was hunting him.'}}

\subsection*{Impersonal constructions}
Constructions with `det är/var' (\emph{`it is/was'}) is very common in Swedish \cite[\textsection 309d]{H&H}:
\enumsentence{
\shortex{6}{Det & var & roligt & att & höra.}
{it &was & nice &to &hear.}
{`I'm glad to hear that.'}}
'Det' (\emph{it}) is also used as formal subject in presentational constructions where
the real subject is put in the position of an object.
\enumsentence{
\shortex{6}{Det &står &en &älg &på &fältet.}
{ it & stands &a & moose &on &the field}
{`There is a moose in the field.'}}

\subsection*{Reflexive pronouns}
\label{swe:refl}
The Scandinavian language have special reflexive pronouns
and possessive pronouns for the 3rd
person\cite[\textsection 310 \& 319]{H&H}. They are distinct from the normal 3rd person forms.
\enumsentence{
\begin{tabular}{llll}
a. & Han slog \textbf{sig}.&\hspace{20mm}b. & Han såg \textbf{sitt} barn.\\
&\emph{He hit him self.} && \emph{He saw his (own) child.}
\end{tabular}}
\enumsentence{
\begin{tabular}{llll}
a. &Han slog \textbf{honom}. &b.& Han såg \textbf{hans} barn.\\
&\emph{He hit him (another person).}&&\emph{He saw his (another person's) child.}
\end{tabular}}
The 1st and 2nd persons use the normal personal pronoun in object form as reflexive
pronouns.

%Mention varandra. Same idea.
%Basic info about Swedish. V2 lang, inverted word order and subordinate clauses.
%Passive, reflexive 'sitt', 'det är', 'det sitter en katt där', prepare the reader
%for what will be written in the grammar part.
%Adjectives but not verbs congruate with nouns. De är stora, They are big, Jag
%är stor, I am big
%Since the s-passive is the most common one in written Swedish \cite{SAG-34-1}, it has

\section{Saldo}
\label{sec:saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built on Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The ~~words are equipped ~~ with semantic, syntactic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

If one word has different meanings, there will be separate entries for each of
them.  The word ``växa" (``grow") for example, has two meanings: \emph{getting
bigger}
\enumsentence{Gräset växte varje dag\\
The grass grew each day}
or \emph{being a plant}. 
\enumsentence{Det växte en gammal ek där.\\
There stood an old oak.}
The word ``växa" is however declined the same way for both interpretations of the word.
Hence Saldo's morphological lexicon, provided under LPGL in XML format,
only keeps one entry of ``växa".
For the homonyms that do not share the whole inflection tables, different entries are 
kept in the morphological lexicon.

@@

\section{Grammatical Framework}
\label{sec:gf}
The main component of the the project is the Grammatical Framework (GF) \cite{gfbok}, %It is
a grammar formalism based on Martin-Löf type theory\cite{martinlof}. GF is
designed for multilingual applications and the formalism is equivalent to
Parallel Multiple Context-Free Grammars (Seki \& al. 1991), it is hence
stronger than mildly contex-free grammars and can express whatever tree
adjoining grammars (Joshni,xx) and HG?? can.  --very confusing here!!

GF is a strongly typed functional programming languages, inspired by
%programming languages like 
ML \cite{ml} and Haskell \cite{haskell}. Its type
checking system is inspired by \textlambda Prolog\cite{bratko} and the
dependent types by Agda\cite{agda}. 
>> check this!!
The first version of the framework was implemented in 19xx in xx and has
since then fatt en trogen skara användare, med headquarter 
at Gothenburg University.

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. >>check!
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF)\ref{pgf},
which can be used with %There are libraries for using PGF files with
Java, Java script, C and Haskell. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to try and test her grammar by using commands for parsing, vizualization of parse trees,
random generation, word alignment, morphological quizzes etc.
%The following example shows how to parse a sentence and visualizing the parse tree.
%, by using the \textit{pipe} \verb-|-, sending the result to 
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser dig" | visualize_parse
> parse "jag ser dig" | visualize_tree
\end{verbatim}
\caption{Example of how the GF shell can be used. The first command will parse the given
sentence and visualize the resulting parse tree and the second will visualize the 
abstract syntax tree.}
\label{fig:shellvp}
\end{figure}
>> test vp -format=pdf -view=evince
>> test vt -format=pdf -view=evince -nofun

The visualization of the trees is done with graphviz\footnote{http://www.graphviz.org/}.
Figure \ref{fig:gftree1} shows the results from running the commands in figure \ref{fig:shellvp}
%a visualization of a abstract syntax tree and a parse tree.
\begin{figure}[h]
>> example, typ : i see you. 
\caption{Visualized GF trees}
\label{fig:gftree1}
\end{figure}
The reader should keep in mind that the visualized parse trees is not a
complete representation.
%>> how parse trees work, not correct complete representation.
%>> not a constituent structure. compare to lfg, fstructure, cstructure.
%>> explain how they are created, graphviz and how a user uses them.
%>> fotnoote that they are not complete

\subsection{Writing a GF grammar}
The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntax. The abstract syntax represents
the internal structure and models the semantics without concern for language
specific features such as agreement or word order.
One abstract grammar can be implemented by a number of concrete grammars, each
representing one language. As a comparison, the abstract and concrete syntax
may be thought of as fstructures and cstructures in LFG (x,x).

@@

>> move much of this to Swedish GF part?
Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|,
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:\\
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. To do this, we add a category \verb|NP| to the
>> reformulate so that I don't sound stupid
abstract. \verb|N| now means a noun
in any of the number, and we introduce two function, \verb|NSg : N -> NP| and
\verb|NPl : N -> NP| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; NP ; V ; S ;               lincat V, S, NP = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : NP -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> NP ;                    NPl n = n.s ! Pl ;
    NPl : N -> NP ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records,tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s| shows that it is a table, which given a parameter of type \verb|Num| returns
a string. \verb|Num| is defined as a parameter, which can either have value \verb|Sg|
or \verb|Pl|. In \verb|NPl| and \verb|NSg|, \verb|n.s| means that we use the
field \verb|s| of \verb|n|, and then the selection operator \verb|!| is used to
select a branch in the table.

@@

>>not very restricted, not few ambiguities.
>>the scope is controlled language, but possible ambiguities (discuss this in future
>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
>>syntactical (man with the telescope), 
>>lexical (youSg/Pl, växa).
A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt},
a project which aims to develop language independent material for mathematics problems,
and the formal software verification tool KeY\cite{key}, in wich GF is used for
translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
>> use talk here and move key to other place??
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
European project MOLTO\cite{molto} for online translation between up to 15 language.\\

\subsection{The resource library}
%>>Best part? Biggest library?
%>>Most general resource, first large grammar, covers fundamental syntactics.
The GF package provides a powerful resource library\cite{gf-resource}, which covers the
fundamental morphology and syntactics for more than 20 languages. %is a very useful part of the framework. It is provided in the 
%, they give a good start for all who want to write an application grammar.
%of a language for each application, all
%this can be found as libraries. So far there are 20 languages in the library, and
The languages all share the same abstract and they constitute  
a very useful resource when implementing application grammars.
There is also a small test lexicon included, containing some hundred of common
words.
